Reasoning Graphically about Quantum Computation

### 2008-03-07 papers/reasoning-graphically/paper.tex
close to final edit - spell check 1
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1408 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

\newcommand{\vinterp}[1]{\lfloor\hspace{-0.27em}|\, #1\, |\hspace{-0.27em}\rfloor}
\newcommand{\binterp}[1]{\lceil\hspace{-0.27em}|\, #1\, |\hspace{-0.27em}\rceil}
\newcommand{\minterp}[1]{\llbracket #1 \rrbracket}
\newcommand{\sinterp}[1]{| #1 |}

@@

\begin{abstract}
  Recent graph-based formalisms of quantum computation provide an
  abstract and symbolic way to represent and simulate
  computations. However, manual manipulation of such graphs is slow
  and error prone. We present a formalism, based on compact closed
  categories, that supports mechanised reasoning about such
  graphs. This gives a compositional account of graph rewriting that
  preserves the underlying categorical semantics. Using this
  representation, we describe a generic system with a fixed logical
  kernel that supports reasoning about models of compact closed
  category. A salient feature of the system is that it provides a
  declarative account of derived results. We illustrate the framework
  by instantiating it for a graphical language of quantum computation
  and show how this can be used to perform symbolic computation.

@@

Recent work in quantum computation has emphasised the use of graphical
languages motivated by the underlying logical structure of quantum
mechanics
itself~\cite{AbrCoe:CatSemQuant:2004,Selinger:dagger:2005,Coecke2005Kindergarten-Qu,Coecke2006POVMs-and-Naima,Coecke2006Quantum-Measure}.
These techniques have a number of advantages over the convential
matrix-based approach to quantum mechanics:

\begin{itemize}
\item The visual representation abstracts over the values in the
  matricies. This removes detail that requires a lot of work for a
  human to interpret.

\item Many properties have a natural graphical representation. For
  example, non entanglement is can be infered from disjoint subgraphs.

\item The graphical calculus generalises to domains other than just
  vector spaces. In particular, it provides a representation for
  compact closed cartegories.

@@

The main contribution of this paper is a graph-vased formalism that is
suitable for representing and evaluating quantum computations. We
start by reviewing a formalism for graphical representations of
compact closed categories. We also revisit a model of quantum
computation based on this calculus. We then extend the graphical
calculus in two significant ways driven by the need to express rules
that could not be accounted for in the initial formalism. By combining
these extensions, we provide a representation that captures an
interesting and useful set of graph patterns. Notably, it can express
the Spider Theorem, which is normally written using informal elipses
notation. We provide a semantics for our graph-based formalism in
terms of the initial representation of compact closed categories.

@@

\noindent The intuition behind this definition is that the subgraph of
$G$ determined by $V'$ should be preserved exactly by $f$; it should
contain $G$ without any extra incident egdes into the verticies $V'$
whereas other vertices may be identified and may contain additional
verticies.

Given two graphs, $G$ and $H$, a pair of partial maps, $r_V: V_G
\rightharpoonup V_H$ and $r_E: E_G \rightharpoonup E_H$, is called a
partial graph morphism if whenever $r_E(e)$ is defined then
$r_V(s_G(e))$ and $r_V(t_G(e))$ are defined, and the restriction of
$r_E$ to its preimage satifies equations \eqref{eq:graph-hom1} and
\eqref{eq:graph-hom2} above.

@@

\begin{definition}
\label{compactcat-def}
A strict symmetric monoidal  category
\cite{MacLane:CatsWM:1971,AspLon:CatTypStruct:1991} is called compact 
closed \cite{KelLap:comcl:1980} when each object $A$ has a chosen dual
object $A^*$, and morphisms
\begin{gather*}
  d_A : I \to A^* \otimes A\\
  e_A : A \otimes A^* \to A
\end{gather*}
such that
\begin{align}
  A \iso A \otimes I \rTo^{\id{A} \otimes d_A} A \otimes A^* \otimes A
  \rTo^{e_A \otimes \id{A}} I \otimes A \iso A & = \id{A} \label{eq:comcl1}\\
  A^* \iso I \otimes A^* \rTo^{ d_A \otimes \id{A^*}} A^* \otimes A
  \otimes A^* \rTo^{\id{A^*} \otimes e_A} A^* \otimes I \iso A^* & =
  \id{A^*} \label{eq:comcl2}
\end{align}
\end{definition}

@@

\begin{proposition}
Concrete graphs form a compact closed category whose objects are
ordered signed sets and whose arrows $f:A\to B$ are concrete graphs with
$\cod f = B$  and $\dom f = A^*$.
\end{proposition}
For each ordered signed set $A$, the identity map for $A$ has
$\dom \id{A} = A^*$ and $\cod \id{A} = A$; its underlying
graph has $E = A$ and $V = A^* + A$ with $t(a) = a$ and $s(a) =
a^*$.  Given a pair of concrete graphs $f:A\to B$ and $g:B\to C$ their
composition $g\circ f:A\to C$ is constructed by merging the two
graphs, erasing the vertices of $\cod f$ and $\dom g$, and identifying
the edges previously incident at the deleted vertices.  (Due to the
opposite polarity of the domain and codomain the edges have compatible
direction.)  The tensor product on objects $A,B$ is simply $A+B$;
given $f:A\to B$, $g: C\to D$, the graph of $f \otimes g$ is the
disjoint union of the graphs of $f$ and $g$.  The unit for the tensor
is the empty set.  The morphisms $d_A : I \to A^* \otimes A$,
$e_: A \otimes A^* \to I$ have the same underlying graph, but $\dom d =
\emptyset$, $\cod d = A^*+A$, $\dom e = A+A^*$ and $\cod e = \emptyset$.

@@

As a final remark before moving on, note that the external structre of vertex
in a concrete graph is esentially the same as that of a complete
graph;  hence one can consistently view subgraphs as vertices, and
abstract over the their internal structure.

@@

  \caption{Graphical Equations for Quantum Systems. The $(\cdot)^\dag$
    functor gives a vertical symmetry to the category, hence for every
    equation we have a second equation obtained by flipping the
    diagram upside down.  In additon, we have a ``colour duality'':
    each equation shown here gives rise to second, which is obtained
    by exchanging the two colours. The colour duality is derivable
    from the equations for involving $H$.}
  \label{fig:graph-quant-eqns}
  \figureend
\end{figure}

@@

\begin{figure}[t]
$$\begin{array}{ccc}
\inlinegraphic{6em}{spider_lhs} & = & \inlinegraphic{6em}{spider_rhs}
\end{array}$$

\label{fig:spider}\caption{An informal equation on graphs
  that expresses the Spider Theorem. }
\end{figure}

Hence any connected subgraph involving nodes of only one colour,
wherever it occurs in a graph, may be collapsed to a single vertex
labelled by a single value $\alpha$, giving a ``spider''. Informally,
this can be depicted graphically as the equation in
Figure~\ref{fig:spider}. Conversely, a spider may be
arbitrarily divided into subspiders, provided the total in- and
out-degree is preserved, along with and the sum of the $\alpha $s.
Further, one can derive, from the spider theorem, $n$-fold versions of
many of the other equations.

@@

\begin{center}
\inlinegraphic{2.5em}{compact1} \quad\quad
\inlinegraphic{2.5em}{compact3} \quad\quad
\inlinegraphic{2.5em}{compact5}
\end{center}

\noindent However, composition with semi-circles ($d$ and $e$ from
Definition~\ref{compactcat-def}), on the co-domain or domain, allows
an equation involving one of the above to easily lead to a derivation
corresponding to either of the other two. For instance, proving any of
the following allows a trivial derivation of the others:

\begin{center}
\inlinegraphic{2.5em}{compact1} $=$
\inlinegraphic{1.3em}{compact2} $\Leftrightarrow$
\inlinegraphic{2.5em}{compact3} $=$ \inlinegraphic{1.3em}{compact4}
$\Leftrightarrow$ \inlinegraphic{2.5em}{compact5} $=$
\inlinegraphic{2.5em}{comonoid-unit4}
\end{center}

\noindent This motivates a characterisation that has only one
representative for such a set of equations. More presicely, we
formalise a representation that abstracts over the boundary nodes
membership in the domain or co-domain. This gives rise to a
\emph{variable-node graphs}, in which boundary nodes have been
generalised to \emph{variable nodes}. The intuition of variable nodes
is that they can replaced by concrete nodes in some graph in a process
analgous to composition.

@@

To formalise the semantics of variable-node graphs, we define
matching, which captures the intuitive idea of a graph with variable
nodes occuring within another graph.

@@

A graph with variable nodes, $G$, can be given a formal semantics by
being interpretated as a set of concrete graphs, denoted by
$\vinterp{G}$. The interpretation is simply the set of concrete graphs
which the variable node graph matches. We refer to members of this set
as instances. Matching is also well-defined between two variable-node
graphs, in which case we denote that $G$ macthes $H$ by $G \leq_v H$.

\begin{theorem}
\label{match-interp-thm}
$G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. The
proof from left to right is a consequence of the trivial lemma that a
compositon of open embeddings is an open embedding. This makes the
embedding of $G$ into $H$ composed with $H$ into $\vinterp{H}$ an open
embedding of $G$ into $\vinterp{H}$. From right to left, we compose
the open embedding $\vinterp{G}$, restricted to it's subset
$\vinterp{H}$, with the inverse of $\vinterp{H}$ to get $G \leq_v H$.
\end{theorem}

@@

@@

A crucial feature of rules like the Spider Theorem is that an arbirary
number of repetitions of a subgraph need to be matched. To support
this we introduce an operation, !-boxing (pronounced bang-boxing), on
graph representations. Given a graph representation, this introduces a
new notation by outlining a set of nodes. These nodes are said to be
in a !-box. Intuitively, the resulting !-boxed graph can be thought of
as representing a set of graphs with an arbirary number of copies of
the !-boxed nodes, where every copy connects, in the same way, to the
nodes outside the !-box. The !-boxes can also be instantiated with
zero copies. This erases all edges to and from the !-box.  The Spider
Theorem in Figure~\ref{fig:spider} for instance is
represented with !-boxes in Figure~\ref{fig:spider-thm-patt}.

@@

More formaly, a !-box graph is a graph paired with a set !-boxes. Each
of these !-boxes is the set of nodes inside it and each !-box must be
disjoint from the others.\footnote{The disjointness condition avoids a
  slightly more complex, although also more expressive, semantics for
  overlaping node sets in the !-boxes. While such expressivity is
  interesting from a representational point of view, it is not needed
  to express the currently known rules for quantum computation.}  We
now define the concept of !-box matching which we then use to provide
a formal semantics for the !-boxed graphs.

\subsubsection{Matching.}

To formalise the intuitive notion that a !-box represents arbitrary
number of copies of the subgraph made from it's nodes, we introduce
\emph{!-box matching}. This is a binary relation on graphs, written
infix as $G \leq_! H$ when $G$ matches $H$. The set matching !-box
graphs is defined as the set closed by the following operations on the
!-boxes in a graph:

@@

\subsubsection{Semantics.}

We give a formal semantics to a !-box graphs by defining the !-box
graph in terms of a set of graphs in the underlying representation. We
denote the interpretation of a !-box graph $G$ by $\binterp{G}$ and
say that members of $\binterp{G}$ are instances of $G$.

@@

\begin{theorem}
\label{thm:bang-box-respect}
\emph{!-Matching respects !-box semantics}: $G \leq_! H
\Leftrightarrow \binterp{G} \supseteq \binterp{H}$. The proof is a
simple consequence from the defintion of $\binterp{G}$ being a subset
of the graphs that match $G$.
\end{theorem}

@@

The representation of Compact Closed Categories as Graphs, discussed
in~\S\ref{sec:graph-repr-comp}, is too restrictive for reasoning about
quantum computation. In particular, graphical rules such the Spider
Theorem, from Figure~\ref{fig:informal-spider-law}, are frequently
needed, but not expressible. In this section we combine the {\em
  !-boxes} and {\em variable-node} extensions to graphs. We call this
representation \emph{graph patterns}. This forms a representation that
allows us to express, in a finite way, the infinite family of concrete
graphs equations. In particular, the Spider Theroem can now be
represented by the equation shown in Figure~\ref{fig:spider-thm-patt}.
We also extend the notion of matching for graph pattens.
%as well as present a corresponding notion of plugging. 
This provides the foundations for the rewriting machinery
in~\S\ref{sec:rewriting} which can then be used to reasoning about
quantum computation.

\begin{figure}[t]
$$\begin{array}{ccc}
\inlinegraphic{12em}{spider_lhs_patt} & \;=\; & \inlinegraphic{9em}{spider_rhs_patt}
\end{array}$$
\label{fig:spider-thm-patt}\caption{The Spider Theorem expressed using
  graph patterns. The !-boxes are named $i$ and $j$. The variable
  nodes are white and named $a$ and $b$. The internal node's data,
  which is an angle, is written inside the node.}
\end{figure}

@@

\subsubsection{Semantics as Concrete Matches.}

The semantics for a graph pattern $G$ is a set of concrete graphs
denoted by $\minterp{G}$ and define it as:

@@

\subsubsection{Graph Pattern Matches.}

The specification for one pattern, $G$, to match another one, $H$ is
that it is more general with respect to the interpretation:
$\minterp{G} \supseteq \minterp{H}$. However, graph patterns can
correspond to a countably infinite number of concrete graphs. Thus
matching between graph pattern cannot be implemented by simply
unfolding them to their interpretated concrete graphs and checking the
membership relation.

Fortunately, it is quite easy to provide a decidable matching. This is
because the size of the unfolding that needs to be considered can be
bounded. The key observation is that a graph $G_1$ will never match a
graph with less non-variable nodes. Thus unfolding of $G_1$ can be
bounded by the number of non-variable nodes in $G_2$. While this gives
a generate and test style algorithm, it is not efficient. The
intuition for an efficient algorithm is to search through one graph
incrementally increasing the matched part and avoids symmetries in
matching. The development of an efficient algotihm is ongoing work.

@@

In this section we describe how the garph pattern formalism can
provide a \emph{meta-level} framework for reasoning about models of
compact closed categories. Following the terminology of logical
frameworks such as Isabelle~\cite{isabelle}, we call the
specification, provided by the underlying model, an
\emph{object-level} graph formalism. An object-level formalisation
defines a set of rules which are treated as the axioms for the
system. It also defines the data at the nodes and edges as well as
corresponding data-matching behaviour. For it's part, the meta-level
provides generic machinery to manipulate graphs and derive new
rules. In this section we describe the meta-level framework and prove
its adequacy for rewriting.

@@

The intuitive idea of substituon with a rule is to replace a subgraph
that matches the left hand side with the right hand side. However, not
all pairs of rules are make sense with respect to the underlying
semantics. For an equation to be well defined with respect to the
compact closed structure it must not be possible to changes the type
of a concrete graph by rewriting. Mapping this restriction back to
pairs of graph patterns results in the following conditions on rules:

@@

\item There has to be a isomorphism between variable nodes in the left
  and right hand subgraphs. More concretely, there must be a unique
  naming of the variable nodes in the left hand side such that each
  name occurs in a unique variable node on the right hand side. Given
  a matching on the left hand side, the target subgraph is replaced
  with the right hand side while keeping the same instantiations for
  the isomorphic variable nodes.

\item Rules must also define a partial 1-1 mapping between !-boxes on
  the left and right hand sides. The intuition for this mapping is
  that the unfolding used when matching a !-box on the left, is
  applied to the mapped !-box on the right before replacement.

\item The interplay between !-boxes and variable nodes means that when
  a variable node appears within a !-box on one side of a rule it must
  also appear under a mapped !-box on the other side.

@@

Notationally, and implementationally, we annotate !-boxes and variable
nodes in a graphs with unique names. For example, see
Figure~\ref{fig:spider-thm-patt} which illustrates the Spider Theorem
as a rule between graph patterns. In this figure, the partial 1-1
mapping for !-boxes is represented by !-boxes having the same name on
the left and right of a rule.  Similarly, the isomorphism between
variable nodes is captured by the set of variable node names on the
left and right hand side being equal.

@@

The axioms of an object system typically come from the semantics of
the unerlying system. For instance, the equations given in
Figure~\ref{fig:graph-quant-eqns} provide a system of reasoning about
quantum computation. In some cases, such equations can be provdied by
the underltying semantics. When such rules are expressed as graph
patterns, we replace the concrete representations boundary nodes with
variable nodes. This operation is called \emph{lifting}. An equation
on graph patterns corresponds to a possibly infinite set of equations
betweens concrete graphs. Thus we might worry that the lifted
equations express too much: they may allow rewrites which are not
intended.

We call the property that the lifted representation is a conservative
extension of the initial theory \emph{adequacy}. For models of compact
closed categories, the proof of adequacy is quite simple: given an
equation between concrete graphs, $G = H$, we observe that every
instance of the lifted equation corresponds to the original equations
composed with some graph. The graph is given by the unmapped subgraph
using the open embedding from the lifted equations onto their
instance. Thus if $G = H$ is true, then so is every instance of it's
lifting, and thus the lifting produces an adequate representation.

@@

Having defined what make a valid rule, we now present the meta-logic
of the framework. This is quite simple as they only involve dealing
with object-level equations:

@@

\noindent where $\Gamma$ is the set of object-level axioms and for the
reflexivity rule, $A$ must be a well-formed pattern graph. We assume
that axioms in $\Gamma$ meet the validity conditions described
earlier. These rules all preserve the validity conditions on equations
and thus the system as a whole ensures only valid rules are derived.

A new graph can be introduced using the reflexivity law (\emph{refl}).
By applying the \emph{subst} rule, intermediate results are derived
which can themselves be used to rewrite other rules and conjetcures.
In this way, the system allows derived rules to provide an
abbreviation for a combination of steps. Sets of rules can be applied
automatically to simulate computation in an the object domain. In
\S\ref{sec:case-study} we illustrate this for quantum computation.

\subsection{Soundness and Completeness}

The soundness of a system that uses our meta-level rests on two
arguments:

\begin{itemize}
\item Representation of the underying model as a compact closed
  category with a defined set of equations.

\item The soundness and adequacy of the axioms when expressed within
  the compact closed category. Adquency ensures that the encoding as a
  graph pattern represents only the intended concrete rules. The
  soundness of the rules is obviously needed and requires proving the
  rule in terms of the underlying semantics.
\end{itemize}

>>>>

%\newcommand{\vinterp}[1]{\lfloor\hspace{-0.27em}|\, #1\, |\hspace{-0.27em}\rfloor}
%\newcommand{\binterp}[1]{\lceil\hspace{-0.27em}|\, #1\, |\hspace{-0.27em}\rceil}
\newcommand{\vinterp}[1]{\llbracket #1 \rrbracket_v}
\newcommand{\binterp}[1]{\llbracket #1 \rrbracket_!}
\newcommand{\minterp}[1]{\llbracket #1 \rrbracket}
\newcommand{\sinterp}[1]{| #1 |}

@@

\begin{abstract}
  Recent graph-based formalisms of quantum computation provide an
  abstract and symbolic way to represent and simulate computations.
  However, manual manipulation of such graphs is slow and error prone.
  We present a formalism, based on compact closed categories, that
  supports mechanised reasoning about such graphs. This gives a
  compositional account of graph rewriting that preserves the
  underlying categorical semantics. Using this representation, we
  describe a generic system with a fixed logical kernel that supports
  reasoning about models of compact closed category. A salient feature
  of the system is that it provides a formal and declarative account
  of derived results that can include `ellipses'-style notation. We
  illustrate the framework by instantiating it for a graphical
  language of quantum computation and show how this can be used to
  perform symbolic computation.

@@

Recent work in quantum computation has emphasised the use of graphical
languages motivated by the underlying logical structure of quantum
mechanics
itself~\cite{AbrCoe:CatSemQuant:2004,Selinger:dagger:2005,Coecke2005Kindergarten-Qu,Coecke2006POVMs-and-Naima,Coecke2006Quantum-Measure}.
These techniques have a number of advantages over the conventional
matrix-based approach to quantum mechanics:

\begin{itemize}
\item The visual representation abstracts over the values in the
  matrices. This removes detail that requires a lot of work for a
  human to interpret.

\item Many properties have a natural graphical representation. For
  example, non entanglement is can be inferred from disjoint subgraphs.

\item The graphical calculus generalises to domains other than just
  vector spaces. In particular, it provides a representation for
  compact closed categories.

@@

The main contribution of this paper is a graph-based formalism that is
suitable for representing and evaluating quantum computations. We
start by reviewing a formalism for graphical representations of
compact closed categories. We also revisit a model of quantum
computation based on this calculus. We then extend the graphical
calculus in two significant ways driven by the need to express rules
that could not be accounted for in the initial formalism. By combining
these extensions, we provide a representation that captures an
interesting and useful set of graph patterns. Notably, it can express
the Spider Theorem which is normally written using informal ellipses
notation (see Figure~\ref{fig:spider}). We provide a semantics for our
graph-based formalism in terms of the initial representation of
compact closed categories.

@@

\noindent The intuition behind this definition is that the subgraph of
$G$ determined by $V'$ should be preserved exactly by $f$; it should
contain $G$ without any extra incident edges into the vertices $V'$
whereas other vertices may be identified and may contain additional
vertices.

Given two graphs, $G$ and $H$, a pair of partial maps, $r_V: V_G
\rightharpoonup V_H$ and $r_E: E_G \rightharpoonup E_H$, is called a
partial graph morphism if whenever $r_E(e)$ is defined then
$r_V(s_G(e))$ and $r_V(t_G(e))$ are defined, and the restriction of
$r_E$ to its preimage satisfies equations \eqref{eq:graph-hom1} and
\eqref{eq:graph-hom2} above.

@@

\begin{definition}
\label{compactcat-def}
A strict symmetric monoidal  category
\cite{MacLane:CatsWM:1971,AspLon:CatTypStruct:1991} is called compact 
closed \cite{KelLap:comcl:1980} when each object $A$ has a chosen dual
object $A^*$, and morphisms
\begin{gather*}
  d_A : I \to A^* \otimes A \quad\quad\quad e_A : A \otimes A^* \to A
\end{gather*}
such that
\begin{align}
  A \iso A \otimes I \rTo^{\id{A} \otimes d_A} A \otimes A^* \otimes A
  \rTo^{e_A \otimes \id{A}} I \otimes A \iso A & = \id{A} \label{eq:comcl1}\\
  A^* \iso I \otimes A^* \rTo^{ d_A \otimes \id{A^*}} A^* \otimes A
  \otimes A^* \rTo^{\id{A^*} \otimes e_A} A^* \otimes I \iso A^* & =
  \id{A^*} \label{eq:comcl2}
\end{align}
\end{definition}

@@

\begin{proposition}
Concrete graphs form a compact closed category whose objects are
ordered signed sets and whose arrows $f:A\to B$ are concrete graphs with
$\cod f = B$  and $\dom f = A^*$.
\end{proposition}
For each ordered signed set $A$, the identity map for $A$ has $\dom
\id{A} = A^*$ and $\cod \id{A} = A$; its underlying graph has $E = A$
and $V = A^* + A$ with $t(a) = a$ and $s(a) = a^*$.  Given a pair of
concrete graphs $f:A\to B$ and $g:B\to C$ their composition $g\circ
f:A\to C$ is constructed by merging the two graphs, erasing the
vertices of $\cod f$ and $\dom g$ (called the \emph{boundary
  vertices}), and identifying the edges previously incident at the
deleted vertices.  (Due to the opposite polarity of the domain and
codomain the edges have compatible direction.)  The tensor product on
objects $A,B$ is simply $A+B$; given $f:A\to B$, $g: C\to D$, the
graph of $f \otimes g$ is the disjoint union of the graphs of $f$ and
$g$.  The unit for the tensor is the empty set.  The morphisms $d_A :
I \to A^* \otimes A$, $e_: A \otimes A^* \to I$ have the same
underlying graph, but $\dom d = \emptyset$, $\cod d = A^*+A$, $\dom e
= A+A^*$ and $\cod e = \emptyset$.

@@

As a final remark before moving on, note that the external structure
of a vertex in a concrete graph is essentially the same as that of a
complete graph; hence one can consistently view subgraphs as vertices,
and abstract over the their internal structure.

@@

  \caption{Graphical Equations for Quantum Systems. The $(\cdot)^\dag$
    functor gives a vertical symmetry to the category, hence for every
    equation we have a second equation obtained by flipping the
    diagram upside down.  In addition, we have a ``colour duality'':
    each equation shown here gives rise to second, which is obtained
    by exchanging the two colours. The colour duality is derivable
    from the equations for involving $H$.}
  \label{fig:graph-quant-eqns}
  \figureend
\end{figure}

@@

\begin{figure}[t]
$$\begin{array}{ccc}
\inlinegraphic{6em}{spider_lhs} & = & \inlinegraphic{6em}{spider_rhs}
\end{array}\qquad
\begin{array}{ccc}
\inlinegraphic{12em}{spider_lhs_patt} & \;=\; & \inlinegraphic{9em}{spider_rhs_patt}
\end{array}$$
\label{fig:spider}\caption{ [Left] An informal equation on graphs that
  expresses the Spider Theorem.  [Right] The Spider Theorem expressed
  formally using graph patterns.  The !-boxes are named $i$ and $j$.
  The variable nodes are white and named $a$ and $b$. The non-variable
  node data (the angle) is written inside the node.}

\end{figure}

Hence any connected subgraph involving nodes of only one colour,
wherever it occurs in a graph, may be collapsed to a single vertex
labelled by a single value $\alpha$, giving a ``spider''. Informally,
this can be depicted graphically as the equation in
Figure~\ref{fig:spider}. Conversely, a spider may be
arbitrarily divided into sub-spiders, provided the total in- and
out-degree is preserved, along with and the sum of the $\alpha $s.
Further, one can derive, from the spider theorem, $n$-fold versions of
many of the other equations.

@@

\begin{center}
\inlinegraphic{2.5em}{compact1} $=$
\inlinegraphic{1.3em}{compact2} $\qquad\quad$
\inlinegraphic{2.5em}{compact3} $=$ \inlinegraphic{1.3em}{compact4}
$\qquad\quad$ \inlinegraphic{2.5em}{compact5} $=$
\inlinegraphic{2.5em}{comonoid-unit4}
\end{center}

%\begin{center}
%\inlinegraphic{2.5em}{compact1} \quad\quad
%\inlinegraphic{2.5em}{compact3} \quad\quad
%\inlinegraphic{2.5em}{compact5}
%\end{center}

\noindent However, composition with semi-circles ($d$ and $e$ from
Definition~\ref{compactcat-def}), on the co-domain or domain, allows
an equation involving one of the above to easily lead to a derivation
corresponding to either of the other two: proving any of the above
allows a trivial derivation of the others.

To address this, we formalise a representation that abstracts over the
boundary nodes membership in the domain or co-domain. This gives rise
to a \emph{variable-node graphs}, in which boundary nodes have been
generalised to \emph{variable nodes}. The intuition of variable nodes
is that they can replaced by concrete nodes in some graph in a process
analogous to composition.

@@

To formalise the semantics of variable-node graphs, we define
matching, which captures the intuitive idea of a graph with variable
nodes occurring within another graph:

@@

A graph with variable nodes, $G$, can be given a formal semantics by
being interpreted as a set of concrete graphs, denoted by
$\vinterp{G}$. The interpretation is simply the set of concrete graphs
which the variable node graph matches. The above definition of
matching also applies between two variable-node graphs, in which case
we denote that $G$ matches $H$ by $G \leq_v H$.

\begin{theorem}
\label{match-interp-thm}
$G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. The
proof from left to right is a consequence of the lemma that a
composition of open embeddings is an open embedding. From left to
right, the embedding of $G \leq_v H$ composed with $\vinterp{H}$ is
thus an open embedding of $G$ into $\vinterp{H}$. Hence $\vinterp{G}
\supseteq \vinterp{H}$. From right to left, we compose the open
embedding $\vinterp{G}$, restricted to it's subset $\vinterp{H}$, with
the inverse of $\vinterp{H}$ to get $G \leq_v H$.
\end{theorem}

@@

These theorems ensure that macthing is what we expect it to be and
also give us the usual properties such as associativity.

@@

A crucial feature of rules like the Spider Theorem is that an arbitrary
number of repetitions of a subgraph need to be matched. To support
this we introduce an operation, !-boxing (pronounced bang-boxing), on
graph representations. Given a graph representation, this introduces a
new notation by outlining a set of nodes. These nodes are said to be
in a !-box. Intuitively, the resulting !-boxed graph can be thought of
as representing a set of graphs with an arbitrary number of copies of
the !-boxed nodes, where every copy connects, in the same way, to the
nodes outside the !-box. The !-boxes can also be instantiated with
zero copies. This erases all edges to and from the !-box.

% The Spider
%Theorem in Figure~\ref{fig:spider} for instance is
%represented with !-boxes in Figure~\ref{fig:spider-thm-patt}.

@@

More formally, a !-box graph is a graph paired with a set !-boxes. Each
of these !-boxes is the set of nodes inside it and each !-box must be
disjoint from the others.\footnote{The disjointness condition avoids a
  slightly more complex, although also more expressive, semantics for
  overlapping node sets in the !-boxes. While such expressivity is
  interesting from a representational point of view, it is not needed
  to express the currently known rules for quantum computation.}  We
now define the concept of !-box matching which we then use to provide
a formal semantics for the !-boxed graphs.

\textbf{Matching.} To formalise the intuitive notion that a !-box
represents arbitrary number of copies of the subgraph made from it's
nodes, we introduce \emph{!-box matching}. This is a binary relation
on graphs, written infix as $G \leq_! H$ when $G$ matches $H$. The set
matching !-box graphs is defined as the set closed by the following
operations on the !-boxes in a graph:

@@

\textbf{Semantics.} We give a formal semantics to a !-box graphs by
defining the !-box graph in terms of a set of graphs in the underlying
representation. We denote the interpretation of a !-box graph $G$ by
$\binterp{G}$ and say that members of $\binterp{G}$ are instances of
$G$.

@@

\begin{theorem}
\label{thm:bang-box-respect}
\emph{!-Matching respects !-box semantics}: $G \leq_! H
\Leftrightarrow \binterp{G} \supseteq \binterp{H}$. The proof is a
simple consequence from the definition of $\binterp{G}$ being a subset
of the graphs that match $G$.
\end{theorem}

@@

The representation of Compact Closed Categories as Graphs, discussed
in~\S\ref{sec:graph-repr-comp}, is too restrictive for reasoning about
quantum computation. In particular, graphical rules such the Spider
Theorem, from Figure~\ref{fig:spider}-left, are frequently needed, but
not expressible. In this section we combine the {\em !-boxes} and {\em
  variable-node} extensions to graphs. We call this representation
\emph{graph patterns}. This forms a representation that allows us to
express, in a finite way, the infinite family of concrete graphs
equations. In particular, the Spider Theorem can now be represented as
shown in Figure~\ref{fig:spider}-right. We also extend the notion of
matching for graph pattens.
%as well as present a corresponding notion of plugging. 
This provides the foundations for the rewriting machinery
in~\S\ref{sec:rewriting} which can then be used to reasoning about
quantum computation.

@@

\textbf{Semantics as Concrete Matches.} The semantics for a graph
pattern $G$ is a set of concrete graphs denoted by $\minterp{G}$ and
define it as:

@@

\textbf{Graph Pattern Matches.} The specification for one pattern,
$G$, to match another one, $H$ is that it is more general with respect
to the interpretation: $\minterp{G} \supseteq \minterp{H}$. However,
graph patterns can correspond to a countably infinite number of
concrete graphs. Thus matching between graph pattern cannot be
implemented by simply unfolding all interpretions as concrete graphs
and checking the membership relation.

Fortunately, it is quite easy to provide decidable matching: the size
of the unfolding that needs to be considered can be bounded. The key
observation is that a graph $G_1$ will never match a graph with less
non-variable nodes. Thus unfolding of $G_1$ can be bounded by the
number of non-variable nodes in $G_2$. While this gives a generate and
test style algorithm, it is not efficient. The intuition for an
efficient algorithm is to search through one graph incrementally
increasing the matched part and avoids symmetries in matching.  
%The development of an efficient algorithm is ongoing work.

@@

In this section we describe how the graph pattern formalism can
provide a \emph{meta-level} framework for reasoning about models of
compact closed categories. Following the terminology of logical
frameworks such as Isabelle~\cite{isabelle}, we call the specification
provided by the underlying model an \emph{object-level} graph
formalism. An object-level formalisation defines a set of rules which
are treated as the axioms for the system. It also defines the data at
the nodes and edges as well as corresponding data-matching behaviour.
For it's part, the meta-level provides generic machinery to manipulate
graphs and derive new rules. We now describe the meta-level framework,
noting the conditions for a rule to be valid, and prove the systems
adequacy for rewriting.

@@

The intuitive idea of substitution with a rule is to replace a
subgraph that matches the left hand side with the right hand side.
However, not all pairs of rules are make sense with respect to the
underlying semantics. For an equation to be well defined with respect
to the compact closed structure it must not be possible to changes the
type (the boundary nodes in the domain and co-domain) of a concrete
graph by rewriting.  Mapping this restriction back to pairs of graph
patterns results in the following conditions on rules:

@@

\item There has to be a isomorphism between variable nodes in the left
  and right hand subgraphs. Given a matching on the left hand side,
  the target subgraph is replaced with the right hand side while
  keeping the same instantiations for the isomorphic variable nodes.

\item Rules must also define a partial bijective mapping between
  !-boxes on the left and right hand sides. The intuition for this
  mapping is that the unfolding used when matching a !-box on the
  left, is applied to the mapped !-box on the right before
  replacement.

\item The interplay between !-boxes and variable nodes means that when
  a variable node appears within a !-box on one side of a rule, it must
  also appear under a mapped !-box on the other side.

@@

Notationally, and implementationally, we annotate !-boxes and variable
nodes in a graphs with unique names. For example,
Figure~\ref{fig:spider-thm-patt} shows the Spider Theorem. In this
figure, the mapping between !-boxes is represented by !-boxes having
the same name on the left and right of a rule.  Similarly, the
isomorphism between variable nodes is captured by the set of variable
node names on the left and right hand side being equal.

@@

The axioms of an object formalism come from the semantics of the
underlying system. For instance, the equations given in
Figure~\ref{fig:graph-quant-eqns} can be proved by matrix
calucluations oi the underlying model. When such rules are expressed
as graph patterns, we replace the concrete representation's boundary
nodes with variable nodes. This operation is called \emph{lifting}. An
equation on graph patterns corresponds to a infinite (when their are
variable nodes) set of equations between concrete graphs. Thus we
might worry that the lifted equations express too much: they may allow
rewrites which are not true.

We call the property that the lifted representation is a conservative
extension of the initial theory \emph{adequacy}. For models of compact
closed categories, the proof of adequacy is quite simple: given an
equation between concrete graphs, $G = H$, we observe that every
instance of the lifted equation corresponds to the original equations
composed with some graph. The graph is given by the unmapped subgraph
using the open embedding from the lifted equations onto the considered
instance. Thus if $G = H$ is true, then so is every instance of it's
lifting, and thus lifting produces an adequate representation.

@@

Having defined what make a valid rule, we now present the meta-logic
of the framework. This is quite simple as it only involves dealing
with object-level equations:

@@

\noindent $\Gamma$ is the set of object-level axioms.

We assume
that axioms in $\Gamma$ meet the validity conditions described
earlier. These rules all preserve the validity conditions on equations
and thus the system as a whole ensures only valid rules are derived.

The reflexivity rule (\emph{refl}) assumes that $A$ is a well-formed
pattern graph. This rule allows a new graph can be introduced. By then
applying the \emph{subst} rule, intermediate results are derived which
can themselves be used to rewrite other rules and conjectures.  In
this way, the system allows derived rules to provide an abbreviation
for a combination of steps.

A sets of rules can be applied automatically to simplify a graph or
simulate computation in an the object domain. For this to terminate, a
suitable left-to-right ordering on rules can be observed such as the
size of the subgraph must decrease. In \S\ref{sec:case-study} we
illustrate simulating a quantum computation.

% \subsection{Soundness and Completeness}

% The soundness of a system that uses our meta-level rests on two
% arguments:

% \begin{itemize}
% \item Representation of the underlying model as a compact closed
%   category with a defined set of equations.
